Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-09-01 ebayes_rnaseq_heterosis.tex
use EH

Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means. We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH). Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, heterosis, and underdominance.  However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.  Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).  For simplicity, throughout the remainder of the article, we will use the term heterosis to describe the situation where the hybrid mean is more extreme than the parental means, i.e. we say there is heterosis if and only if either HPH or LPH holds.

@@

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for heterosis.

@@

In the maize context that motivates this work, we are interested in heterosis, either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here. \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid effect ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e. $\hat{\delta}_g$ and $\hat{\alpha}_g$. For example,  
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study. Section \ref{s:maize} describes the maize dataset in detail. We compared our method to approaches using the R packages \edgeR{} , \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.

@@

For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$. Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have heterosis. For many heterosis genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$. Thus there are many genes in these simulated data sets that are considered to have heterosis, but whose signal in the simulations will be extremely small. Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many heterosis genes. Therefore, we expect it will be difficult to accurately identify heterosis genes, but believe this level of difficulty is representative of real applications.

@@

In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the heterosis context. For each method, we attempted to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for heterosis.

\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for heterosis.

@@

For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with heterosis, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without heterosis. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Example ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
\label{f:roc}
\end{figure}
The ROC curves indicate modest performance, e.g. for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%. This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the low signal-to-noise ratio in these simulated data.

For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

@@

With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}. Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of heterosis is slightly different due to INLA not providing a full posterior.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a ridge at an effect size of zero for probabilities below 0.5. Above a probability of 0.5, we see a prototypical volcano pattern, with increased heterosis probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high heterosis probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of heterosis and large estimated effect sizes are candidates for further investigation.

>>>>

Depending on whether large or small values of a phenotype are favorable, hybrid vigor can occur if the mean phenotype of a hybrid is greater than both parental means or less than both parental means. We refer to the former as high-parent heterosis (HPH) and the latter as low-parent heterosis (LPH). Note that heterosis, HPH, and LPH are similar to the quantitative genetics concepts of dominance, heterosis, and underdominance.  However, the various forms of dominance are usually reserved for describing the association of mean phenotype with homozygous and heterozygous genotypes at a single genetic locus.  Heterosis involves a comparison of inbreds (simultaneously homozygous at many loci) and hybrids (simultaneously heterozygous at many loci).  For simplicity, throughout the remainder of the article, we will use the term \emph{extreme heterosis} to describe the situation where the hybrid mean is more extreme than the parental means, i.e. we say there is heterosis if and only if either HPH or LPH holds.

@@

To use \RNAseq{} counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure, and calculate empirical Bayes posterior probabilities for extreme heterosis.

@@

In the maize context that motivates this work, we are interested in extreme heterosis (EH), either low-parent heterosis (LPH) or high-parent heterosis (HPH), in gene expression. 
For a specific gene $g$, LPH occurs when average expression in the hybrid is less than the average expression of either parent, i.e. $\delta_g < -|\alpha_g|$ or, equivalently, $\mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}$, and HPH occurs when average expression in the hybrid is greater than the average expression of either parent, i.e. $\delta_g > |\alpha_g|$ or, equivalently, $\mu_{g3} > \max\{\mu_{g1},\mu_{g2}\}$. 
% and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
% \begin{align}
% H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
% H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
% \label{e:hypotheses}
% \end{align}
% where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
% We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
% \begin{align}
% P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
% &\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
% \end{align}
% where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. 
We evaluate these probabilities based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{equation}
%P(LPH_g|y) \approx 
P(\mbox{LPH}_g|y,\hat{\pi},\hat{\gamma}) = 
%P\left(\left. \mu_{g3} < \min\{\mu_{g1},\mu_{g2}\}\right|y,\hat{\pi},\hat{\gamma}\right) = 
P\left(\left.\delta_g < -|\alpha_g|\right|y, \hat{\pi},\hat{\gamma}\right) = \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\delta_g^{(m)} < -\left|\alpha_g^{(m)}\right|\right), 
\label{e:probs}
\end{equation}
where $\left(\delta_g^{(m)},\alpha_g^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. HPH probability is defined similarly with the inequality reversed and without the negative sign. We construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

@@

In addition to the approach above, we utilized \ShrinkBayes{} to estimate EH probabilities with two modifications described here. \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid effect ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate EH probabilities conditional on posterior means for the half-parental difference and hybrid effect, i.e. $\hat{\delta}_g$ and $\hat{\alpha}_g$. For example,  
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \le 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

To assess the efficacy of our method to identify genes demonstrating EH, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study. Section \ref{s:maize} describes the maize dataset in detail. We compared our method to approaches using the R packages \edgeR{} , \baySeq{} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and \ShrinkBayes{} \citep{van2014shrinkbayes}.

@@

For a particular gene, the truth was determined via the estimated values for $\alpha_g$ and $\delta_g$. Specifically, if $|\hat{\delta}_g| >|\hat{\alpha}_g|$ the gene was considered to have EH. For many heterosis genes, the value of $|\hat{\delta}_g|$ was only slightly larger than $|\hat{\alpha}_g|$. Thus there are many genes in these simulated data sets that are considered to have EH, but whose signal in the simulations will be extremely small. Conversely, there are many non-heterosis genes whose value of $|\hat{\delta}_g|$ was only slightly smaller than $|\hat{\alpha}_g|$, but whose simulated data will look similar to many EH genes. Therefore, we expect it will be difficult to accurately identify EH genes, but believe this level of difficulty is representative of real applications.

@@

In addition to the approach of \citeauthor{ji2014estimation}, we modified two existing \RNAseq{} approaches, \edgeR{} and \baySeq{}, for use in the heterosis context. For each method, we attempted to provide a measure of the strength of EH for each gene such that large values of this measure indicate support for EH.

\edgeR{} can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of EH, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using \edgeR{}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we defined a new $p$-value denoted as $p_g$ and set to $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant significance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no EH for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for EH.

\baySeq{} allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of \edgeR{}, we used the posterior probabilities of these hypotheses to construct a measure of EH for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using \edgeR{}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the two hypotheses that allow for EH.

@@

For the methods in Sections \ref{s:method} and \ref{s:alternative}, we sorted genes according to the computed measure of the strength of evidence for EH. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to distinguish genes with EH, as defined in Section \ref{s:sim_data}, 
% From Dan, "why this citation?"
from those without EH. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Example ROC curves for false positive rates below 0.1 for the approaches using \citeauthor{ji2014estimation}, \edgeR{}, and \baySeq{} described in Section \ref{s:alternative}, the \ShrinkBayes{} approach described in Section \ref{s:shrinkbayes} and the eBayes approach described in Section \ref{s:ebayes} using both normal and laplace distributions for the half-parental difference and hybrid effect.}
\label{f:roc}
\end{figure}
The ROC curves indicate modest performance, e.g. for a false positive rate of 5\%, the best performing methods only achieved a true positive rate of just over 15\%. This is consistent with our expectation discussed in Section \ref{s:sim_data} due to the low signal-to-noise ratio in these simulated data.

For this simulation, we can see that the approaches based on the model in Section \ref{s:model}, i.e. eBayes and \ShrinkBayes{}, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with EH from genes without improves.

@@

With 4 replicates per variety, there does not appear to be much of a difference between \ShrinkBayes{} and the eBayes approaches, but as the number of replicates increases, the eBayes approaches appear to improve relative to \ShrinkBayes{}. Two difference exist between the \ShrinkBayes{} and eBayes approaches that could explain the difference in AUC: 1) \ShrinkBayes{} utilizes a different empirical Bayes approach for estimating both the hyperparameters and the gene-specific parameters and 2) the measure of EH is slightly different due to INLA not providing a full posterior.

@@

Figure \ref{f:volcano} provides a volcano plot, in this case a bivariate histogram, to visualize the maximum of the probabilities of LPH and HPH versus estimated effect size. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{volcano}}
\caption{A bivariate histogram of the maximum of the LPH and HPH probabilities versus estimated effect size defined in equation \eqref{e:effect_size} for the B73 $\times$ Mo17 maize experiment.}
\label{f:volcano}
\end{figure}
The figure shows a ridge at an effect size of zero for probabilities below 0.5. Above a probability of 0.5, we see a prototypical volcano pattern, with increased EH probability corresponding to larger estimated effect sizes and no estimated effect sizes near zero for genes with high EH probability. We also see asymmetry, with larger negative effect sizes than positive effect sizes due to genes with hybrid counts near zero and relatively high parental counts. Genes with high estimated posterior probabilities of EH and large estimated effect sizes are candidates for further investigation.
